Web Processing - Standardised GIS Analyses for Cable Route Planning

### 2023-03-11 paper/discussion.tex
style, grammar format, figures

% !TeX spellcheck = en_US
As shown, the need for computation time increases with the power of two with the resolution. 
Similarly, the usage of main memory increases. 
Which in tern limits the processable data points and resolution.
On the error only scales linear with the resolution. 
Hence, there is a diminishing return of smaller errors, compared to compute time and resources used.

Therefore, this worked tries a) to reduce the needed to compute power and b) decrease the deviation for a given resolution.

For b), two methods have been used to increase the similarity of the paths computed from medium resolution raster, to the path of the highest resolution raster.
These methods can be used as surrogates for the more complex calculation of the least cost path with the higher resolution.
In the first method a bi-linear down sampling of the higher resolution raster has been applied and in the second method, the all touched false and true raster where averaged in different ratios, to compute the optimum mixed raster, for the used costs.
While the second method of using an averaged raster, a higher similarity to the path from the highest resolution raster, the down sampling method is more
simple and does not need to be optimized for the given costs.
Hence, it is easier and widely applicable.

For a), to reduce the computational complexity, the calculation of the medium resolution raster has been applied, to reduce the search space of the aggregation.
While this method reduces the calculation time and usage of main memory, the backtracking part stays unchanged.

As the original least cost path QGIS plug-in the used algorithm stops the least aggregation of the costs after the final end point has been found.
This early stopping might result in suboptimal paths around the end points for some edge cases, where the connection via another neighbor might be more optimal.

The set of rules that are used create the cost raster, includes a rule for creating a buffer around buildings  which is set to the level \textit{prohibited} areas.
The resolution of the medium level raster needs to be high enough to crudely show every detail, Hence at least in the magnitude of the minimum object size plus twice its buffer, in this example in the range of at least 25~m or 10~m to include the streets in all touched false raster.
Other details as rivers and houses, ae already includes in lower resolution raster, because of the buffering.
Although the overall result can change, due super position of objects and there buffer.
Nevertheless, does higher resolution resemble, the geography more precise, but in addition show paths that are too small, to be taken.
On the other hand, as the run from the starting point to the end point at a resolution of 2.5~m shows, new ways, that could not be reached before, do hint inconsistencies of the rules that construct the cost raster.

Object and there buffer might leave a  path that, is smaller than 10~m resolution.
So a second rule to find a good resolution, that it should be in the magnitude of the size of the structure  to be build.
While the used data set for buildings, also includes power poles, the area of the level \textit{preferential} power grid, includes isles of high costs. 
In further works, these should be excluded.

In this work, the effect of computation costs and deviation of the results, is examined for a very limited set of points.
Also, only the costs of finding the least cost path from one starting point, to a single end point has been considered.
When, multiple endpoints are used, the computation cost for the aggregated cost raster has only to computed once.
On the other hand the effect early stopping is not as dramatic, because the algorithm would only stop early for the last end point with the highest cost.
When calculating several paths from one raster, the benefits of chapter~\ref{subsec:faster-processing-of-the-cost-path-algorithm} is reduced.
Especially, the pre-calculation on lower resolution raster and buffering of the resulting paths as a restricted search area decreases in its effectivity.
As the number of paths rises, fewer of points in higher resolution cost raster will be deselected for the final estimation of the least cost path.
The least cost algorithm does only select the most cost-effective path.
Therefore, paths of similar, but slightly worse cost are stay unknown.
An end-user might be interested in selecting a path from a set of possible paths of similar aggregated costs and apply an adapted set evaluation criteria.
This can be achieved a by adapting the backtracking, to get an array of paths or by applying perturbation on the costs.

In this work, the single cost raster layers aggregated with the maximum function. 
Another possible aggregation function is the sum.
Each aggregation function can be justified, by a different interpretation of the costs and its scale.
The reasoning of selection the maximum function is, limiting the maximum costs.
When the \textit{prohibited} level is used as the highest level.
Then higher levels are \textit{prohibited} within the concept.
This an area even more \textit{prohibited}, when it is included in two different rules of the level?
On the other side, This aggregation is unable to differentiate between these nuances, as  the sum function is.
This way, one can differential between, different sublevels.
One consequence is, that a small mediation between high and low single cost raster, in the final cost raster is possible.

The fact, that all touched False raster lead to more similar results compared to high resolution raster, is due to the fact, that the default level is relatively low.
In cause the default level would rise, the effect probably would flip for low resolution raster.
For high resolution raster, this effect would still hold true, because the fact, that the raster pixel center is used for sampling, reflects the original geometry better.

Interpretation: Interpretation the Cost Paths are much nearer smaller, than 100~m!
No problem because buffering is included in the cost raster.
Not algorithmic speed-ups as pre-compilation, JIT, vector units, multi-core, weight range

>>>>

% !TeX spellcheck = en_GB
In theory, the need for computing time increases with the resolution as power of two. 
Similarly, the use of main memory increases. 
This again limits the number data points, that can be processed and resolution, and probably causes difference in increase of computation time from power of 2 to a power of 2.2, because additional slower ram moduls have been used for the aggregation of higher resolution.

On the other hand, the similarity to higher resolutions only scales linear with the resolution. 
Thus, there is a diminishing return of smaller errors, compared to compute time and resources used.

Therefore, this work attempts a) to reduce the computation power needed and b) reduce the deviation for a given resolution compared to a higher resolution raster.

For a), to reduce the computational complexity, clipping of the high resolution has been applied, to reduce the search space of the aggregation.
While this method reduces the computation time and memory usage, the backtracking part stays unchanged.

For b), two methods have been used to increase the similarity of the paths computed from the medium resolution raster, to the path of the highest resolution raster.
These methods are used as surrogates for the more complex calculation of the Least Cost Path with the higher resolution.
In the first method a bi-linear downsampling of the higher resolution raster has been applied and in the second method, the all touched false and true raster where averaged in different ratios, to compute the optimum weighted cost raster.
While the second method of using an averaged raster, a higher similarity to the path from the highest resolution raster, the downsampling method is simpler and does not need to be optimised for the given cost.
This disadvantage could be reduced by normalising the costs.

This early stopping may result in suboptimal paths around the end points for some edge cases, where the connection via another neighbour might be more optimal.

The set of rules that are used create the cost raster, includes a rule to create buffer around buildings which is set to the level \textit{prohibited} areas.
In all touched false rasters.
The resolution of the medium level raster needs to be high enough to show every detail.
At least in the magnitude of the minimum object size plus twice its buffer.
This is true for the 10~m resolution raster and less true for the 25~m resolution raster that misses some details for roads for all touched false raster.
Other details such as rivers and houses, are already included in the lower resolution raster, due to larger buffers.
The Least Cost Path algorithm searches for an optimal path as a line.
As lines do not have a width, the route found might contain bottlenecks, that have a smaller width than the object that should be placed there.
Therefore, the used resolution should not be smaller than, the width of the object that should be placed.
This can be avoided by downsampling, but by weighting the medium resolution all touched true or false rasters.

This paper examines the effect of computational costs and deviation of the results, is examined for a very limited set of points.
Also, only the cost of finding the Least Cost Path from a single start point, to a single end point has been considered.
If multiple endpoints are used, the computational cost for the aggregated cost raster has only to be paid once.

If multiple paths are calculated from a single raster, the speed-up benefit is reduced.
Especially, pre-calculation on medium resolution raster and clipping around a buffering of the resulting medium resolution paths becomes less effective.
As the number of paths increases, fewer pixels are clipped.
The Least Cost Path algorithm does only select the single most cost-effective path.
Therefore, paths of similar, but slightly higher costs remain unknown.
In additional, slight variations on the costs rasters can lead to very different paths, although the costs will not change much.
An end-user may be interested in selecting a path from a set of similar aggregated costs and applying their own evaluation criteria.
This can be achieved a by adjusting the backtracking and return polygons, or by applying perturbation on the costs.

In this work, the intermediate cost raster layers are aggregated using the maximum function. 
Another possible aggregation function is the sum or average.
Each aggregation function can be justified, by a different interpretation of the cost and its scale.

When the \textit{prohibited} level is used as the highest level, then summing the two highest levels would result in a new highest level. 
Also, the maximum function does not interfere with the nodata value. 
Although this can be done by a nansum- / nanmean-function, if the nodata value is set to \textit{not a number} during the aggregation.
The disadvantage of aggregation with the maximum function is, that this aggregation is unable to distinguish between nuances of different overlapping intermediate costs.
On sum or average aggregated rasters, one can distinguish between, different sublevels.

The fact that all touched false rasters produce more similar results than high resolution raster, is probably due to the fact, that the default level is relatively low.
As the default level increases, the effect would probably be reduced for low resolution rasters.
For high resolution raster, the effect would still be present, because the fact, that the raster pixel centre is used for sampling, reflects the original geometry better.

This effect of the similar aggregated costs per resolution could also be seen in the test paths, even when the paths varied greatly with the change in resolution or algorithm.
This could be an indication of an even spatial distribution of the costs.

The all touched true cost raster shows every detail, but the sampling with all touched true increases the size of the features.
The fact that the aggregated costs per resolution for all touched true rasters overestimate the costs when computing the path from a low resolution raster, might be due to the fact, that the values of the costs are not uniformly distributed, but that the high costs are much more frequent, because the costs are exponential scaled.

### 2023-03-11 paper/methods.tex
grammar + change to en_GB
style, grammar format, figures

The nationwide land coverage (ATKIS) with a scale of 1:250000 are provided by the Federal Agency for Cartography
and Geodesy~\cite{noauthor_digitales_2021}.
The national power grid (tags: 'power': line) has been retrieved via OpenStreetMap~\cite{boeing_osmnx_2017}.
Local data as houses at Level of Detail 1 are provided by the State Office for Geoinformation and Land Surveying of
Lower Saxony~\cite{noauthor_opengeodatani_2022}.
In addition, local planning geodata for the land use are taken
from 'Metropolplaner' (Planning data Lower Saxony \& Bremen)~\cite{noauthor_metropolplaner_2022}.

@@

All these layers are provided as vectors.
The Least Cost Path algorithm uses raster data.
Rasterisation transforms a vector into a raster.
Rasterisation can be done in two different ways.
In both ways, the rasterisation can be imaged, as the old vector is superimposed on the new raster grid with the new
given resolution and the new affine transformation and the coordinate reference system of the vector.
Both rasterisation techniques differ in the selection of the pixel, that describe the original polygon.
The pixels can be selected , if either the centre of that pixel is overlapped
by the geo-object, or any part of the pixel is overlaid.
The version with any part overlapped is called all touched True.
The version where a overlapped pixel centre is require is called all touched False.
All touched False is considered the default (see figure~\ref{fig:alltouched}).
\begin{figure}[!ht]
	\centering

	\resizebox{8cm}{!}{
		\begin{tikzpicture}
			\node[draw] at (0,6.5) {\huge all touched false};
			\draw[step=1cm,gray,very thin] (-1.9,-1.9) grid (5.9,5.9);
			\draw[fill=blue, thick] (-0.2, 0.2)--(-0.6, 3.1)--(-0.2,4.1)--(1.2,3.4)--(2.8,3.6)--(3.2, 2.6)--(3.4, 2.1)--(1.8,1.3)--(1.2, 0.1)--cycle;

			\draw[ultra thick,->, red] (6.5, 2) --node[above=1mm] {\huge rasterization} (8.5, 2);

			\draw[step=1cm,gray,very thin] (9.1,-1.9) grid (16.9,5.9);
			\draw[fill=blue] (11,0)--(11,2)--(10,2)--(10,3)--(11,3)--(11,4)--(12,4)--(12,3)--(13,3)--(13,4)--(14,4)--(14,2)--(13,2)--(13,1)--(12,1)--(12,0)--cycle;
		\end{tikzpicture}%
	}%
	\vspace*{5mm}
	\resizebox{8cm}{!}{
		\begin{tikzpicture}
			\node[draw] at (0,6.5) {\huge all touched true};
			\draw[step=1cm,gray,very thin] (-1.9,-1.9) grid (5.9,5.9);
			\draw[fill=blue, thick] (-0.2, 0.2)--(-0.6, 3.1)--(-0.2,4.1)--(1.2,3.4)--(2.8,3.6)--(3.2, 2.6)--(3.4, 2.1)--(1.8,1.3)--(1.2, 0.1)--cycle;

			\draw[ultra thick,->, red] (6.5, 2) --node[above=1mm] {\huge rasterization} (8.5, 2);

>>>>

The national land coverage (ATKIS) with a scale of 1:250000 are provided by the Federal Agency for Cartography
and Geodesy~\cite{noauthor_digitales_2021}.
The national power grid (tags: 'power': line) has been retrieved via OpenStreetMap~\cite{boeing_osmnx_2017}.
Local data as houses at Level of Detail 1 are provided by the State Office for Geoinformation and Land Surveying of
Lower Saxony~\cite{noauthor_opengeodatani_2022}.
In addition, local planning geodata for the land use are taken
from 'Metropolplaner' (Planning data Lower Saxony \& Bremen)~\cite{noauthor_metropolplaner_2022}.

@@

All these layers are provided as vectors.
The Least Cost Path algorithm uses raster data.
Rasterisation transforms a vector into a raster.
Rasterisation can be done in two different ways.
In both ways, the rasterisation can be imaged, as the old vector is superimposed on the new raster grid with the new
given resolution and the new affine transformation and the coordinate reference system of the vector.
Both rasterisation techniques differ in the selection of the pixel, that describe the original polygon.
The pixels can be selected , if either the centre of that pixel is overlapped
by the geo-object, or any part of the pixel is overlaid.
The version with any part overlapped is called all touched True.
The version where an overlapped pixel centre is require is called all touched False.
All touched False is considered the default (see figure~\ref{fig:alltouched}).
\begin{figure}[!ht]
	\centering

	\resizebox{8cm}{!}{
		\begin{tikzpicture}
			\node[draw] at (0,6.5) {\Huge all\_touched=false};
			\draw[step=1cm,gray,very thin] (-1.9,-1.9) grid (5.9,5.9);
			\draw[fill=blue, thick] (-0.2, 0.2)--(-0.6, 3.1)--(-0.3,4.1)--(1.2,3.4)--(2.8,3.6)--(3.2, 2.6)--(3.4, 2.1)--(1.8,1.3)--(1.2, 0.1)--cycle;

			\draw[ultra thick,->, red] (6.5, 2) --node[above=1mm] {\Huge rasterisation} (8.5, 2);

			\draw[step=1cm,gray,very thin] (9.1,-1.9) grid (16.9,5.9);
			\draw[fill=blue] (11,0)--(11,2)--(10,2)--(10,3)--(11,3)--(11,4)--(12,4)--(12,3)--(13,3)--(13,4)--(14,4)--(14,2)--(13,2)--(13,1)--(12,1)--(12,0)--cycle;
		\end{tikzpicture}%
	}%
	\vspace*{5mm}
	\resizebox{8cm}{!}{
		\begin{tikzpicture}
			\node[draw] at (0,6.5) {\Huge all\_touched=true};
			\draw[step=1cm,gray,very thin] (-1.9,-1.9) grid (5.9,5.9);
			\draw[fill=blue, thick] (-0.2, 0.2)--(-0.6, 3.1)--(-0.3,4.1)--(1.2,3.4)--(2.8,3.6)--(3.2, 2.6)--(3.4, 2.1)--(1.8,1.3)--(1.2, 0.1)--cycle;

			\draw[ultra thick,->, red] (6.5, 2) --node[above=1mm] {\Huge rasterisation} (8.5, 2);

### 2023-03-11 paper/paper-CableRoute.tex
style, grammar format, figures

\newacronym{wfs}{wfs}{Web Feature Service}
\newacronym{wps}{wps}{Web Processing Service}
\newacronym{res}{res}{resolution}
\newacronym{r}{r}{ratio}
\newacronym{al}{al}{all touched}
\newacronym{t}{t}{True}
\newacronym{f}{f}{False}
\newacronym{agg.}{agg.}{aggregated}
\newacronym{rel}{rel}{relative}
\newacronym{corr}{corr}{corrected}
\newacronym{MCDA}{MCDA}{multi-criteria decision analysis}

@@

	\received{20 February 2007}
	\received[revised]{12 March 2009}
	\received[accepted]{5 June 2009}

@@

	Other methods such as the \acrfull{MCDA}  consider further aspects as stakeholder, experts, public and decision maker~\cite{bertsch_participatory_2016} (see section~\ref{sec:related-works}).

@@

	Older models are e.g.\ are optimizing the well fare, economic modelling is not sufficient for planning modern routes, as the aspect of environmental sustainability, security of supply and the public acceptance more and more play an important role~\cite{hauff_gesellschaftliche_2011} in modern planning.

	In contrast to a GIS analysis, these further factors can be studied done with a \acrfull{MCDA}~\cite{bertsch_participatory_2016}.

	Stakeholder as decision maker can be included and combined with an expert system~\cite{bertsch_participatory_2016}.

	The uncertainties connected with \acrshort{MCDA} data uncertainties, preferential uncertainties and model uncertainties are examined by a with a sensitivity analysis~\cite{bertsch_participatory_2016} or simulation~\cite{butler_simulation_1997}.
	Both inter and intra criteria preferential uncertainties~\cite{bertsch_sensitivity_2007} can be considered.

	The different models of the power system, that are used in the expert system span from not modelling the grid at all to models that includes Kirchhoffs laws~\cite{bertsch_participatory_2016}.
	The calculation of a power system is complex, as changing one edge will change the flow in all parts of grid and the actual grid data are confidential~\cite{bertsch_participatory_2016}.
	Some models include not only the grid of a nation alone, but consider neighbouring states or the overall European power system~\cite{dietrich_will_2010}.

	\section{Conclusion}\label{sec:conclusion}
	The cost of computation of the least cost path scales with the square of the resolution.
	The difference between the aggregated costs of the paths per resolution computed from the all touched false raster and the all touched true raster only shows a linear decrease.
	Hence, the gain in precision per compute time decreases.
	The presented methods tried to circumvent this bottleneck.
	When using the researched strategies, on medium resolution raster, the computation for high resolution results might successfully be reduced, without compromising the run time for worse distances to the least cost path from the high resolution raster.
	The paths might change very strongly even with small changes in total costs.
	Hence to develop an alternative backtracking algorithm that, that generates a set of possible paths within a corridor of costs might be a good strategy, to offer alternative routes and show variations between these routes at the same time.
	An alternative method for future methods, would be to specify a range of costs and compute the aggregation of the costs and superimpose these aggregations and compute the path by backtracking from the superimposed cost raster. 
	This in turn increases the computation time again.

	In an actual search for least cost path a survey might be used as method to estimate the costs.
	When the sample size of the survey is large enough, the
	weights might take local differences relative acceptance into consideration.

	Only changes in the algorithm have been applied in the search for a speed-up.
	There a still other methods as just-in-time compilation.

>>>>

\newacronym{wfs}{wfs}{Web Feature Service}
\newacronym{wps}{wps}{Web Processing Service}
\newacronym{MCDA}{MCDA}{multi-criteria decision analysis}

@@

	\received{14 March 2023}
	%%\received[revised]{12 March 2009}
	%%\received[accepted]{5 June 2009}

@@

	Other methods such as \acrfull{MCDA} consider further aspects as stakeholder, experts, public and decision maker~\cite{bertsch_participatory_2016} (see section~\ref{sec:related-works}).

@@

	The different models of the power system, that are used in the expert system range from not modelling the grid at all to models that include Kirchhoffs laws~\cite{bertsch_participatory_2016}.
	The calculation of a power system is complex, because changing one edge changes the flow in all parts of grid and the actual grid data are confidential~\cite{bertsch_participatory_2016}.
	Some models consider not only the grid of one nation, but consider neighbouring states or the whole European power system~\cite{dietrich_will_2010}.	
	Older national power grid models are e.g.\ are optimising the well fare, but economic modelling is not sufficient for planning modern routes, as the aspect of environmental sustainability, security of supply and the public acceptance play an increasingly important role~\cite{hauff_gesellschaftliche_2011} in modern planning.

	In contrast to a GIS analysis, these other factors can be studied done with a \acrfull{MCDA}~\cite{bertsch_participatory_2016}.
	Stakeholders as decision makers can be included and combined with an expert system~\cite{bertsch_participatory_2016}.

	The uncertainties associated with \acrshort{MCDA} data uncertainties, preferential uncertainties and model uncertainties are investigated by a with a sensitivity analysis~\cite{bertsch_participatory_2016} or simulation~\cite{butler_simulation_1997}.
	Both inter- and intra-criteria preferential uncertainties~\cite{bertsch_sensitivity_2007} can be considered.

	\section{Conclusion}\label{sec:conclusion}
	The cost of computing of the Least Cost Path scales with the square of the resolution.
	The difference between the aggregated costs of the paths per resolution computed from the all touched false raster and the all touched true raster only shows a linear decrease.
	Therefore, the gain in accuracy per compute time decreases.
	The presented methods have attempted to circumvent this bottleneck.
	When these strategies are applied to a medium resolution raster, the compute time for high resolution results can successfully be reduced, without compromising the run time for worse distances to the least cost path from the high resolution raster.
	The paths can change significantly even for a small change in total cost.
	Therefore, to develop an alternative backtracking algorithm, that generates a corridor of costs might be a good strategy, to offer alternative paths and at the same time show variations between these paths.
	An alternative for future methods, would be to specify a range of costs and compute the aggregation of the costs and superimpose these aggregations and compute the path by backtracking from the superimposed cost raster. 
	This again increases the computation time.

	In an actual search for Least Cost Path search, a survey could be used as a method to estimate the costs.
	If the sample size of the survey is large enough, the
	weights could take  into account local differences in relative acceptance.

	Only changes in the algorithm have been applied in the search for acceleration.
	Therefore, other methods such as just-in-time compilation have not been tested.
